\section{Exercises}
The exercises are split into four sets, corresponding to the four exercise sessions held as part of the course. Many of the exercises are collected from previous iterations of the course, and these in turn may originate from \cite{Mun}. A few have been inspired by exercises in \cite{DM}.

\subsection{Set \#1}
\begin{enumerate}[label=1.\arabic*]
  \item Define a relation on $\bbR$ by
    \[
      C=\{(x,y) \mid x - y \in \bbZ \}.
    \]
    Show that $C$ is an equivalence relation and describe the set of equivalence classes of $C$.
  \item Describe all possible topologies on the set $X = \{a,b,c\}$.
  \item Let $X$ be a set, and let $\calT_1$ and $\calT_2$ be two different topologies on $X$. When is the identity map $\id : X \to X$ given by $\id(x) = x$ a continuous map from $(X,\calT_1)$ to $(X,\calT_2)$?
  \item
		Show that the subspace topology $\calT_Y$ is the smallest (meaning coarsest) topology on $Y\subset X$ for which the inclusion $\iota:Y \rightarrow X$ is a continuous map.
	
	\item \label{opens-in-opens} Let $Y\subset X$ be an open (closed) subset of a topological space $X$. Show that a set $U \subset Y$ is open (closed) in the subspace topology on $Y$ if and only if $U$ is open (closed) in $X$.
	
	\item \label{universal-inclusion} Prove Lemma~\ref{universal-inclusion-lemma}.
	
  \item \begin{itemize}
		\item[($a$)] Describe the open sets in the poset topology on $\{a,b,c,d\}$ defined by the relations $a\preceq b\preceq c$ and $a\preceq d$.
		\item[($b$)] Describe the open sets in the poset topology on $(\bbR,\leq)$.
	\end{itemize}
	
  \item The Euclidean space $\bbR^2$ can be identified with the Cartesian product $\bbR \times \bbR$. Use Lemma~\ref{compare-bases} to show that the standard topology on $\bbR^2$ equals the product topology on $\bbR \times \bbR$ (where each $\bbR$ has the standard topology).
  

  \item \label{metric-Hausdorff} Show that metric spaces are always Hausdorff.
  
  \item \label{subspace-Hausdorff} Show that if $X$ is Hausdorff, then so is any subset $Y\subset X$ with the subspace topology.
  
  \item \label{products-Hausdorff} Show that the product of two Hausdorff spaces is Hausdorff.
  
  \item Show that a topological space $X$ is Hausdorff if and only if the diagonal
  \[
    \Delta = \{(x,x) \in X \times X \mid x \in X \} \subset X \times X
  \]
  is closed in the product topology on $X \times X$.
  
  \item \label{metric-first-countable} Let $(X,d)$ be a metric space, and let
  \[
    \calB = \{ B_d(x,1/n) \mid x \in X, n \in \bbN \}.
  \]
  Show that $\calB$ is a basis for the metric topology on $X$.
  
  \item Let $(Y,\preceq)$ be a totally ordered set made into a topological space with the order topology.
  \begin{itemize}
    \item[($a$)] Show that for any two distinct points $x, y \in Y$ there are disjoint neighbourhoods, $U$ and $V$, of $x$ and $y$ respectively, so that $u < v$ for all $u \in U, v \in V$. Conclude that $Y$ is Hausdorff.
    \item[($b$)] Let $X$ be any topological space, and let $f,g:X\to Y$ be two continuous functions. Show that the set $\{x \mid f(x)\preceq g(x)\}$ is closed in $X$.
  \end{itemize}
  
  \item \label{product-metric}Let $(X_1,d_1)$ and $(X_2,d_2)$ be metric spaces. Define a metric on $X_1 \times X_2$ by
	\[
	  d((x_1,x_2),(y_1,y_2)) = \max(d_1(x_1,y_1),d_2(x_2,y_2)).
  \]
  Show that the metric topology on $X_1 \times X_2$ induced by $d$ is the product topology, where $X_1$ and $X_2$ have the metric topologies from $d_1$ and $d_2$ respectively.

  
  \item Let $X,Y,Z$ be topological spaces and consider a function $F:X\times Y\rightarrow Z$. We say that $F$ is \emph{continuous in each variable} if for each $y_0\in Y$ the function $h:X\rightarrow Z$ defined by $h(x)=F(x,y_0)$ is continuous, \emph{and} if for each $x_0\in X$ the function $g:Y\rightarrow Z$ defined by $g(y) = F(x_0,y)$ is continuous. Show that if $F$ is continuous, then $F$ is continuous in each variable.
  
  \item \begin{itemize}
		\item[($a$)] A poset topology is $T_0$. When is it $T_1$?
		\item[($b$)] If $X$ is a $T_0$-space with finitely many elements. Then we can define a relation
		\[x\preceq y \Leftrightarrow y\in \bigcap_{U\subset X\text{ open}, \,x\in U} U.\]
		Show that $\preceq$ is a partial order. What is the poset topology on $(X,\preceq)$?
	\end{itemize}
\end{enumerate}

\subsection{Set \#2}
\begin{enumerate}[label=2.\arabic*]
  \item Show that a topological space $X$ is connected if and only if the following condition holds: if $X = C \cup D$ where $C$ and $D$ are disjoint closed subsets of $X$, then either $C = \emptyset$ or $D = \emptyset$.
  \item Let $X = \{a,b,c\}$ with the topology $\{\emptyset,\{a,b\},\{c\},\{a,b,c\}\}$. Is $X$ connected? Path-connected?
  \item Consider the topology on $\bbR$ generated by the basis $\{(a,\infty) \mid a \in \bbR \}$, and let $x_0 \in \bbR$. \begin{itemize}
    \item[($a$)] What is $\{x_0\}'$, the set of limit points of $\{x_0\}$?
    \item[($b$)] What is the closure $\bar{\{x_0\}}$?
    \item[($c$)] Is $\bbR$ Hausdorff in this topology?
  \end{itemize}
  \item \label{connected-implies-interval} Show that the connected subsets of $\bbR$ are exactly the intervals.
  \item \label{origin-removed-connected} Show that $\bbR^n \setminus \{0\}$ is connected when $n \geq 2$.
  \item \label{invariance-of-dimension-exercise} Show that $\bbR^n \not\simeq \bbR$ when $n > 1$.
  \item Show that the concatenation $\gamma_1 \star \gamma_2$ is continuous by using Theorem~\ref{sequential-continuity} instead of the pasting lemma.
  \item Show that $S^n$ is path-connected for every $n > 0$.
  \item Let $p : X \to Y$ be a quotient map. Show that if $X$ is locally connected then so is $Y$.
  \item \label{locally-connected-open} Show that the connected components of a locally connected space are open.
  \item Let $\{A_n\}_{n \in \bbN}$ be a family of connected subspaces of a topological space $X$ so that $A_n \cap A_{n+1} \not= \emptyset$ for every $n \in \bbN$. Show that $\bigcup_{n \in \bbN} A_n$ is connected.
  \item A space is called \emph{totally disconnected} if its only connected subspaces are one-point sets. Show that if $X$ has the discrete topology, then $X$ is totally disconnected. Does the converse hold?
  \item Let $f : S^1 \to \bbR$ be continuous. Show that there is a point $x \in S^1$ so that $f(x) = f(-x)$.
  \item Let $f : [0,1] \to [0,1]$ be continuous. Show that $f$ has a fixed point, i.e. that there is a point $x \in [0,1]$ so that $f(x) = x$.
  \item \label{exercise-products-connected} Show Theorem~\ref{products-connected} in the case where $I$ is infinite. Inspiration can be found in \cite[Ex.~23.7]{Mun}.
\end{enumerate}

\subsection{Set \#3}
  \begin{enumerate}[label=3.\arabic*]
    \item \label{one-point-exercise}Let $X$ be a topological space.
    \begin{itemize}
      \item[(a)] Show that if $K_1, \dots, K_n$ are compact subspaces, then $K_1 \cup \dots \cup K_n$ is compact.
      \item[(b)] Suppose that $X$ is Hausdorff. Show that if $\{K_i\}_{i \in I}$ is a family of compact subspaces of $X$, then $\bigcap_{i \in I} K_i$ is compact.
      \item[(c)] Prove Proposition~\ref{one-point}.
    \end{itemize}
    \item \label{compact-hausdorff-are-normal-exercise} Show that compact Hausdorff spaces are normal.
    \item \label{local-compactness-in-hausdorff-exercise}Prove Theorem~\ref{local-compactness-in-hausdorff} (Hint: Use the one-point compactification.)
    \item \label{one-point-compactification-unique-exercise}Prove Proposition~\ref{one-point-compactification-unique}.
    \item Show that if $X$ is $T_3$ and $C \subset X$ a closed subset, then the quotient space $X/C$ is Hausdorff.
    \item \label{second-countable-exercise}Show that $\{B(x,r) \mid x \in \bbQ^n, r \in \bbQ_{> 0}\}$ is a basis for the standard topology on $\bbR^n$. Conclude that $\bbR^n$ is second-countable.
    \item Let $Y$ be a compact space, and let $X$ be any topological space.
      \begin{itemize}
      \item[(a)] Show the canonical projection map $\pi : X \times Y \to X$ is closed, i.e. that images of closed sets are closed.
      \item[(b)] Suppose moreover that $Y$ is Hausdorff, and let $f : X \to Y$ be a map. Show that $f$ is continuous if and only if its graph
        \[
          G_f = \{(x,f(x)) \mid x \in X\} \subset X \times Y
        \]
        is closed.
      \end{itemize}
    \item Let $(X,\preceq)$ be a totally ordered set with the order topology, and assume that all every closed interval $[a,b]$ is compact. Show that $X$ has the \emph{least-upper-bound property}\index{least-upper-bound property}; that is, show that every non-empty subset of $X$ which is bounded from above has a least upper bound in $X$.
    \item \label{baires-exercise} Let $X$ be a locally compact Hausdorff space, and let $U_1, U_2, U_3, \dots$ be open dense subsets of $X$. Show that $\bigcap_{n \in \bbN} U_n$ is dense; a result known as the Baire category theorem\index{Baire category theorem}. Hints:
      \begin{itemize}
        \item[(a)] Let $B_0$ be any non-empty open subset of $X$. Construct open sets $B_1, B_2, \dots$ so that
          \[
            \bar{B_n} \subset U_n \cap B_{n-1}
          \]
          for $n \geq 1$ and so that $\bar{B_n}$ is compact for all $n$.
        \item[(b)] Let $K_1, K_2, \dots$ be non-empty compact subsets of a topological space. Show that if we have inclusions $K_1 \supset K_2 \supset \cdots$, then
          \[
            \bigcap_{n \in \bbN} K_n \not= \emptyset.
          \]
          This result is known as Cantor's intersection theorem\index{Cantor's intersection theorem}.
      \end{itemize}
      What happens if we replace the countable family $\{U_n\}_{n \in \bbN}$ with an arbitrary family of open dense subsets?
    \item Let $X$ be a locally compact Hausdorff space, and let $\{F_n\}_{n \in \bbN}$ be a countable family of closed subsets of $X$. Show that if $\Int(F_m) = \emptyset$ for every $m \in \bbN$, then $\Int\left(\bigcup_{n \in \bbN} F_n\right) = \emptyset$. (Hint: use Exercise~\ref{baires-exercise}.)
%     \item Let $X$ be a sequentially compact topological space.
%       \begin{itemize}
%         \item[(a)] If $f : X \to Y$ is continuous, does it follow that $f(X)$ is sequentially compact?
%         \item[(b)] If $C \subset X$ is closed, does it follow that $C$ is sequentially compact?
%         \item[(c)] If $X$ is a subspace of a Hausdorff space $Z$, does it follow that $X$ is closed in $Z$?
%       \end{itemize}
%   \item Let $(X,d)$ be a metric space. A function $f : X \to X$ satisfying
%     \[
%       d(f(x),f(y)) = d(x,y)  
%     \]
%     for all $x, y \in X$ is called an \word{isometry}{isometri} of $X$. Show that if $f$ is an isometry, and $X$ is compact, then $f$ is a homeomorphism. Hint: if $a \in f(X)$, choose $\eps > 0$ so that $B_d(a,\eps) \cap f(X) = \emptyset$. Construct recursively a sequence $\{x_n\}$ so that $x_1 = a$, $x_{n+1} = f(x_n)$ and show that $d(x_m,x_n) \geq \eps$ for $n \not= m$.
%   \item Show that the one-point compactification of $\bbN \subset \bbR$ is homeomorphic to
%     \[
%       \{0 \} \cup \{1/n \mid n \in \bbN \} \subset \bbR.
%     \]
  \end{enumerate}
  
\subsection{Set \#4}
\begin{enumerate}[label=4.\arabic*]
  \item \label{products-of-manifolds-exercise} Prove Lemma~\ref{products-of-manifolds-lemma}.
  \item \label{invariance-of-dimension-manifolds} Use Theorem~\ref{invariance-of-dimension} to show that if $X$ is both an $n$-manifold and an $m$-manifold, then $m = n$.
\end{enumerate}
